---
title: "Introduction to R and Rstudio"
subtitle: "Session2: T-tests and Correlations"
author: "SB Galvin"
output:html_document

---



First I am going to set some options:
```{r}
options(scipen = 999) # Disable scientific notation
```

# T-Tests
## Section 1 Importing or Reading Data

In this case we are going to manually input some data to explain the difference between wide format and long format data

### Wide Formated data:
In wide formatted datasets, one of the variables that constitutes the unit of analysis is mixed with a variable that holds values.In the case of a t test we are interested in differences between two groups, the MALE or FEMALE variables are combined with the response variable 

```{r}
df.wide   <- data.frame(Male = c(01,
                                 02, 02, 
                                 03, 03, 03, 
                                 04, 04, 04, 04,
                                 05, 05, 05, 05, 05, 
                                 06, 06, 06, 06, 06, 06,
                                 07, 07, 07, 07, 07, 07, 07, 
                                 08, 08, 08, 08, 08, 08,
                                 09, 09, 09, 09, 09,
                                 10, 10, 10, 10, 
                                 11, 11, 11, 
                                 12, 12,
                                 13),
                        Female = c(04, 
                                   05, 05,
                                   06, 06, 06, 
                                   07, 07, 07, 07, 
                                   08, 08, 08, 08, 08,
                                   09, 09, 09, 09, 09, 09, 
                                   10, 10, 10, 10, 10, 10, 10,
                                   11, 11, 11, 11, 11, 11, 
                                   12, 12, 12, 12, 12,
                                   13, 13, 13, 13, 
                                   14, 14, 14, 
                                   15, 15, 
                                   16))
df.wide
```

---
###Long Formated data
In long format each row is an observation, and each column is a variable: in this case we have a Response Variable and a condition (SEX) variable:
```{r}
# Create long format data frame
df.long   <- data.frame(Response  = c(01,
                                      02, 02, 
                                      03, 03, 03, 
                                      04, 04, 04, 04,
                                      05, 05, 05, 05, 05, 
                                      06, 06, 06, 06, 06, 06,
                                      07, 07, 07, 07, 07, 07, 07, 
                                      08, 08, 08, 08, 08, 08,
                                      09, 09, 09, 09, 09,
                                      10, 10, 10, 10, 
                                      11, 11, 11, 
                                      12, 12,
                                      13,
                                      04, 
                                      05, 05,
                                      06, 06, 06, 
                                      07, 07, 07, 07, 
                                      08, 08, 08, 08, 08,
                                      09, 09, 09, 09, 09, 09, 
                                      10, 10, 10, 10, 10, 10, 10,
                                      11, 11, 11, 11, 11, 11, 
                                      12, 12, 12, 12, 12,
                                      13, 13, 13, 13, 
                                      14, 14, 14, 
                                      15, 15, 
                                      16),
                        Condition = c("M",
                                      "M", "M",
                                      "M", "M", "M",
                                      "M", "M", "M", "M",
                                      "M", "M", "M", "M", "M",
                                      "M", "M", "M", "M", "M", "M",
                                      "M", "M", "M", "M", "M", "M", "M",
                                      "M", "M", "M", "M", "M", "M",
                                      "M", "M", "M", "M", "M",
                                      "M", "M", "M", "M",
                                      "M", "M", "M",
                                      "M", "M",
                                      "M",
                                      "F",
                                      "F", "F",
                                      "F", "F", "F",
                                      "F", "F", "F", "F",
                                      "F", "F", "F", "F", "F",
                                      "F", "F", "F", "F", "F", "F",
                                      "F", "F", "F", "F", "F", "F", "F",
                                      "F", "F", "F", "F", "F", "F",
                                      "F", "F", "F", "F", "F",
                                      "F", "F", "F", "F",
                                      "F", "F", "F",
                                      "F", "F",
                                      "F"))
df.long
```
If we were to read in a .csv file we could use the `read.csv()` function.

##Section2: Summarise data
Now that we have our data loaded, we can summarise it:

###R base Summary
We can use the `summarise()` 

For long format, we can use `summary()`in conjunction with `by()` to get a quick summary of the data. This tells R to summarise the Dataframe by category membership of Condition variable
```{r}
# Long format
# by() allows you to perform functions divided by a categorical variable
by(df.long,                                      # Tell R where the data is
   df.long$Condition,                            # Tell R what/where the category is
   summary)                                      # Which function you are applying 
```

Wide format data frame. Summary, summarises all columns in a wide data frame
```{r}
# Wide Format
summary(df.wide)
```

###Summary with Psych Package


###Summary with summarytools package
```{r, results = 'asis'}
library(summarytools)
dfSummary(df.long)
dfS
```


##Section 3: Visual Summaries

For df.long only

For both plot summaries I have included the method for saving as a .pdf file, this is commented out to prevent you from writing and overwriting.
The function `pdf()` opens the save call, the last line in the chunk `dev.off()` closes the save call by turning off the graphic device in use

If you wanted to visualise each plot by its self you could copy the code for that plot into another code chunk or R script
```{r}
#pdf("Session2Materials/PlotsS2/S2_Exercises_Panel_Plot.pdf")

# This sets the image parameters, 2 rows, 2 columns
par(mfrow=c(2,2))                               

# Histogram of All responses
hist(df.long$Response,                            # Select variable
     main="Histogram : All Response",             # Set main title
     xlab="Response Variable",                    # Set x axis title
     col = "white",                               # Set Colour
     border = "black"                            # Set bar edge colour
)

# Box plot comparing responses
boxplot(formula = Response ~ Condition,          # Select Y and X variables
        data = df.long,                          # Tell R to use this data 
        main = "Boxplot : Condition x Response", # Set the main title
        col = c("chocolate", "darkmagenta"))     # Set colours

# Histogram of Male responses
hist(df.long$Response[df.long$Condition=="M"],    # Select variable
     main="Histogram : Male Response",            # Set main title
     xlab="Response Variable",                    # Set x axis title
     col = "darkmagenta",                         # Set Colour
     border = "white",                            # Set bar edge colour
     breaks = 14                                  # Set n breaks
)

# Histogram of Female Responses
hist(df.long$Response[df.long$Condition=="F"],    # Select variable
     main="Histogram : Female Response",          # Select main title
     xlab="Response Variable",                    # Set x axis title
     col="chocolate",                             # Set Colour
     border="white",                              # Set bar edge colour
     breaks = 14
)

par(mfrow=c(1,1))                                 # This returns the image parameters to normal
#dev.off()                                         # This closes the graphics device for saving
```


All data appear to be normally distributed. However, we should really check for normality and equal variances before moving on to a t-test

##Section 3: Assumption testing:
### Normality
The t-test assumes a normal distribution. Fortunately we can test this fairly quickly using the `shapiro.test()` function. 

```{r}
shapiro.test(df.long$Response)
```

```{r}
shapiro.test(df.long$Response[df.long$Condition=="M"])
```

```{r}
shapiro.test(df.long$Response[df.long$Condition=="F"])
```

### Equality of variances
```{r}
var.test(Response ~ Condition, data = df.long)
```

```{r}
bartlett.test(Response ~ Condition, data = df.long)
```

## Section 4 T-test
```{r}
# t-test
t.test(Response ~ Condition, 
       data = df.long,
       var.equal=TRUE,
       alternate="greater",
       conf.level = .95)
```

### Effect size
First we can install and load the `effzise` package
```{r}
#install.packages("effsize")
library(effsize)
```

Now use the `cohen.d()` function from `effsize`:
```{r}
cohen.d(Response ~ Condition, data = df.long)
```

### Post-Hoc Power

So heres a user defined function we can use to estimate our tests power
 We need to extract the means and standard deviations from the summary phase of our workflow
```{r}
# Apply mean()
by(df.long$Response,              # Tell R where the data is
   df.long$Condition,             # Tell R what/where the category is
   mean) ->> Mns                  # Which function you are applying # Save to output

# Apply sd()
by(df.long$Response,              # Tell R where the data is
   df.long$Condition,             # Tell R what/where the category is
   sd) ->> StDevs                 # Which function you are applying # Save to output

# This will print out both the means and StDev
c(Mns, StDevs)                    # Print Means First, StDev Second
```

Run this code chunk to use the t-test power function `t.power()`
```{r}
# t.power for equal variances between groups
t.power = function(nsamp=c(10,10),nsim=1000,means=c(0,0),sds=c(1,1),var.equal=TRUE){
    lower = qt(.025,df=sum(nsamp) - 2)
    upper = qt(.975,df=sum(nsamp) - 2)
    ts = replicate(nsim,
       t.test(rnorm(nsamp[1],mean=means[1],sd=sds[1]),
              rnorm(nsamp[2],mean=means[2],sd=sds[2]))$statistic)

    sum(ts < lower | ts > upper) / nsim
}

```

Now insert the n, means and sf for `Condition=="F`, and `Condition=="M`
```{r}
t.power(nsamp=c(49,49),             # N in each sample group
        means=c(10,7),              # Means for each sample group
        sds=c(2.857738, 2.857738))  # StDev for each sample group
```
0.998 % Power, not too bad for a makey uppy data set

If the group variances are unequal, we can set `var.equal=FALSE` in the function call:
```{r}
t.power(nsamp=c(49,49),             # N in each sample group
        means=c(10,7),              # Means for each sample group
        var.equal = FALSE)         # Variances Not Equal
```

Now we can see how well power would recover under this effect size with differig samples. As we have **HIGH** power at 99.8%, and an effect size of **LARGE** 1.05, we can start with a sample size smaller than our current one in the `samps` object:
```{r}
samps = c(10, 15, 20, 30, 100,200,300,400,500)
res = sapply(samps,function(n)t.power(means=c(10,7),nsamp=c(n,n), sds=c(2.586, 2.586)))
names(res) = samps
res
```

So if the effect is true than we may be able to recover it with sample size of 15

We can also use the pwr package to take into account effect size
```{r}
#install.packages("pwr")
library(pwr)
```

Now we can use the pwr.t.test() function. use the helper function to see what arguments the function takes and then fill it 
```{r}
pwr.t.test(n =49, d=1.05, sig.level = 0.05, type="two.sample", alternative = "greater" )
```

NOTES FOR PRESENTER:
Power Function
https://statistics.berkeley.edu/computing/r-t-tests

# Correlation
Correlations are a little bit easier to perform

Correlation functions are very rapid in R
```{r}
NC  <- read.csv("Session2Materials/DataS2/NicCage.csv", fileEncoding="UTF-8-BOM")
NC$Year
```
```{r}
# Histogram of Nic Cage Films
pdf("Session2Materials/PlotsS2/S2_Exercises_HistNicCage.pdf")
hist(NC$Nicholas_Cage,    # Select variable
     main="Histogram : Nic Cage Films",           # Set main title
     xlab="Number of Films",                      # Set x axis title
     col = "tomato",                              # Set Colour
     border = "white",                            # Set bar edge colour
     breaks = 14                                  # Set n breaks
)
dev.off
```

```{r}
# Histogram of Drownings in the US
pdf("Session2Materials/PlotsS2/S2_Exercises_histDrown.pdf")
hist(NC$Drownings,    # Select variable
     main="Histogram : Drownings (US)",           # Set main title
     xlab="Drownings",                            # Set x axis title
     col = "dodgerblue2",                              # Set Colour
     border = "white",                            # Set bar edge colour
     breaks = 14                                  # Set n breaks
)
dev.off()
```

```{r}
plot(NC$Nicholas_Cage, NC$Drownings,
     pch = 21, cex=2, col="grey10", bg="tomato")

```

```{r}
plot(NC$Year, NC$Nicholas_Cage,
     pch = 21, cex=2, col="grey10", bg="tomato",                    # Set point colours and values
     main = "Number of Nicholas Cage Films Released in Each Year",  # Main Title
     xlab = "Year(1999: 2009)",                                     # X axis label
     ylab = "Number of Nicholas Cage Films")                        # Y axis label
par(new = T)                                                        # Use to Add on the lines
lines(NC$Year, NC$Nicholas_Cage)                                    # Add lines
```
```{r}
plot(NC$Year, NC$Drownings,
     pch = 21, cex=2, col="grey10", bg="dodgerblue2",             # Set point colours and values
     main = "Number of Deaths by Drowning in the United States",  # Main Title
     xlab = "Year(1999: 2009)",                                   # X axis label
     ylab = "Number of Drownings (US)")                           # Y axis label
par(new = TRUE)                                                   # Use to Add on the lines
lines(NC$Year, NC$Drownings)                                      # Add lines
```
We can also plot all of the above together
```{r}
#pdf("Session2Materials/PlotsS2/S2_Exercises_NicCage.pdf")
with(NC,plot(Year, Nicholas_Cage,
     pch = 21, cex=2, col="grey10", bg="tomato",                    # Set point colours and values
     main = "Number of Nicholas Cage Films Released in Each Year",  # Main Title
     xlab = "Year(1999: 2009)",                                     # X axis label
     ylab = "Number of Nicholas Cage Films")                       # Y axis label
     )                                
par(new=T)
lines(NC$Year, NC$Nicholas_Cage)                                    # Add lines                                     
par(new = T)
with(NC, plot(Year, Drownings,
              pch = 21, cex=2, col="grey10", bg="dodgerblue2",
               axes=F, xlab=NA, ylab=NA),
     lines(Year, Drownings))

legend("topleft",legend=c("Nicholas Cage Films", "Drownings"),
       lty=c(1,0), pch=c(21, 21),col=c("tomato", "dodgerblue2"))
axis(side = 4)
mtext(side = 4, line = 3, 'Number of Drownings (US)')
par(new = TRUE)                                                   # Use to Add on the lines
lines(NC$Year, NC$Drownings) 
#dev.off()

```

```{r}
cor.test(NC$Nicholas_Cage, NC$Drownings)
```

```{r}
cor.test(NC$Nicholas_Cage, NC$Drownings)->CORRR
c("Var Explained: ",(CORRR$estimate ^2 *100))
```

